---
title: "Introductory R Analysis for Undergraduates"
output:
  html_document:
    number_sections: yes
    html_math: true
    theme: white
    highlight: tango
    code_folding: show
---

```{r setup, include=FALSE}
# Minimal setup - only load essential libraries
knitr::opts_chunk$set(
  comment = "#>",
  eval = TRUE,
  message = FALSE
)

# Load only necessary packages
library(dplyr)
library(ggplot2)  # For data manipulation and plotting
library(afex)     # For ANOVA and post-hoc tests


# Create sample data for demonstrations
set.seed(123)

# Factorial design data (Condition x Trial)
data <- data.frame(
  Condition = rep(c("A", "B"), each = 20),
  Trial = rep(1:4, each = 5, 2)
)
data$Response <- c( rnorm(nrow(data)/2, mean = 50, sd = 10), 
                    rnorm(nrow(data)/2, mean = 45, sd = 10)  )

write.csv(data, "data.csv", row.names = FALSE)

```

---

## 1. Working with Data from CSV Files

This section covers basic data manipulation operations: reading, binding, merging, aggregating, and saving.

### 1.1 Reading CSV Files

```{r read_csv_example}
# Read a CSV file (assuming you have data.csv in your working directory)
df <- read.csv("data.csv")

# View the first few rows
head(df)

# Check data types
str(df)
```

### 1.2 Binding Multiple Data Frames

Combine data from different sources or files.

```{r bind_dataframes}
# Create sample data frames
df1 <- data.frame(ID = 1:5, Value = c(10, 20, 30, 40, 50))
df2 <- data.frame(ID = 6:10, Value = c(60, 70, 80, 90, 100))

# Horizontal binding (side-by-side)
df_horizontal <- cbind(df1, df2)

df_horizontal

# Vertical binding (stacked)
df_vertical <- rbind(df1, df2)

df_vertical

# Using bind_rows() (modern approach)
df_stacked <- bind_rows(df1, df2)

df_stacked
```

### 1.3 Merging and Indexing Data Frames

Combine data based on common columns (like IDs).

```{r merge_dataframes}

# Create sample data with IDs
df1 <- data.frame(ID = 1:5, Value1 = rnorm(5))
df2 <- data.frame(ID = 1:5, Value2 = rnorm(5))

# Merge operation (inner join - only matching IDs)
df_merged <- merge(df1, df2, by = "ID")

df_merged

# we can select a column by name (or by index):
df1$ID

# or by using indexing, where we specify which rows AND which columns we want:
df1[1:5, 1]  # first 5 rows, first column

# we can select rows by index (although sometimes also by name)
# and get the whole row by leaving the column index empty:
df1[3,]

df2[c(2:5),]

# Full outer merge (all IDs from both data frames)
df_full <- merge(df1[c(1:4),], df2[c(2:5),], by = "ID", all = TRUE)

df_full
```

### 1.4 Aggregating Data

Summarize data by grouping variables.

```{r aggregate_data}
# Sample data with grouping variable
data <- expand.grid(
  Group = factor(rep(c("Group1", "Group2"), each = 20)),
  Trial = rep(1:4, 40)
)
data$Value <- rnorm(nrow(data), mean = 50, sd = 10)

# this is a data frame with 6400 rows:
str(data)

# Using base R aggregate()
agg_base <- aggregate(Value ~ Group + Trial, data = data, FUN = mean)

# now we have 8 rows: 2 groups x 4 trials:
agg_base

# the value of Value for Group1 and Trial 1
# is the mean of all Value entries in data
# that have Group == "Group1" and Trial == 1:

# Using dplyr group_by() and summary()
agg_dplyr <- data %>%
  group_by(Group, Trial) %>%
  summarise(Mean = mean(Value, na.rm = TRUE))

agg_plyr

# Using dplyr group_by() and summarise() with multiple statistics
agg_comprehensive <- data %>%
  group_by(Group, Trial) %>%
  summarise(
    Mean = mean(Value, na.rm = TRUE),
    SD = sd(Value, na.rm = TRUE),
    N = n()
  )

agg_comprehensive
```

### 1.5 Saving Data

Save your processed data to files.

```{r save_data}
# Save to CSV
write.csv(df, "processed_data.csv", row.names = FALSE)

# Save with additional parameters
write.csv(df, "data_with_special_chars.csv", 
          row.names = FALSE, na = "NA")

# Save as RDS (R-specific format, preserves all data types)
saveRDS(df, "data.rds")

# Verify files are created
file.exists("processed_data.csv")
```

---

## 2. Creating Standard Plots with ggplot2

This section demonstrates common plot types for visualizing experimental data.

### 2.1 Scatter Plots

```{r scatter_plot}
# Create sample data with different conditions
data <- expand.grid(
  Condition = factor(rep(c("Low", "High"), each = 20)),
  Trial = rep(1:4, 40),
  Response = rnorm(nrow(data), mean = 50, sd = 10)
)

# Basic scatter plot with ggplot2
p_scatter <- ggplot(data, aes(x = Trial, y = Response)) +
  geom_point(aes(colour = Condition), alpha = 0.7) +
  facet_wrap(~Condition) +
  labs(title = "Response Across Trials by Condition",
       x = "Trial Number",
       y = "Response Value")

print(p_scatter)
```

### 2.2 Bar Charts for Comparing Conditions

```{r bar_chart}
# Create summary data for bar chart
summary_data <- data %>%
  group_by(Condition) %>%
  summarise(Mean = mean(Response, na.rm = TRUE),
            SD = sd(Response, na.rm = TRUE)) %>%
  mutate(Mean = as.numeric(Mean))

# Bar chart with error bars
p_bar <- ggplot(summary_data, aes(x = Condition, y = Mean)) +
  geom_bar(stat = "identity", fill = "steelblue") +
  geom_errorbar(aes(ymin = Mean - 1.96 * SD, ymax = Mean + 1.96 * SD),
                width = 0.2, size = 1) +
  labs(title = "Mean Response by Condition",
       x = "Condition",
       y = "Mean Response")

print(p_bar)
```

### 2.3 Line Graphs to Compare Conditions Over Time

```{r line_graph}
# Line plot showing trends over trials
p_line <- ggplot(data, aes(x = Trial, y = Response, colour = Condition, group = Condition)) +
  geom_line() +
  geom_point() +
  labs(title = "Response Trajectory Over Trials by Condition",
       x = "Trial Number",
       y = "Response Value")

print(p_line)
```

---

## 3. Simple Statistical Analysis

This section covers basic statistical tests using base R and the `afex` package.

### 3.1 One-Sample t-Test (Base R)

Test if the mean differs significantly from a specified value.

```{r one_sample_t_test}
# Sample data
data <- data.frame(
  Value = c(45, 48, 52, 55, 58, 60)
)

# Perform one-sample t-test against a hypothesized mean
t_result <- t.test(data$Value, mu = 50)

# Display results
t_result

# Interpretation: If p < 0.05, reject the null hypothesis that the mean is 50
```

### 3.2 Two-Sample t-Test (Base R)

Compare means between two independent groups.

```{r two_sample_t_test}
# Sample data with two groups
data <- expand.grid(
  Group = factor(rep(c("Control", "Treatment"), each = 30)),
  Value = c(rnorm(30, 45, 10), rnorm(30, 55, 10))
)

# Two-sample t-test
t_result <- t.test(data$Value ~ data$Group)

# Display results
t_result

# Use Welch's t-test by default (more robust for unequal variances)
```

### 3.3 ANOVA Using `afex` Package

```{r anova_afex}
# Create a balanced factorial design
data <- expand.grid(
  Factor1 = factor(rep(c("Group1", "Group2"), each = 20)),
  Factor2 = factor(rep(c("LevelA", "LevelB"), each = 20)),
  Value = rnorm(400, mean = 50, sd = 10)
)

# Perform ANOVA using afex package
anova_result <- aov(Value ~ Factor1 * Factor2, data = data)

# Extract ANOVA table
anova_summary <- summary(anova_result)
anova_summary

# Perform post-hoc pairwise comparisons with Tukey's HSD
pairwise_comp <- pairwise.t.test(data$Value, data$Factor1, 
                                 p.adjust.method = "tukey")
pairwise_comp

# Using `anova()` with formula interface
anova_formula <- aov(Value ~ Factor1 * Factor2, data = data)
anova_result <- anova(anova_formula)
anova_result
```

---

```{r wrap_up}
# This notebook demonstrates foundational R skills for undergraduate analysis.
# All code is self-contained and can be adapted to various experimental designs.

# Remember:
# - Always check your data first (head(), str(), summary())
# - Use meaningful variable names
# - Document your analysis steps clearly
# - Use appropriate statistical tests for your research questions

```