---
title: "R Quick Start"
output:
  html_document:
    number_sections: no
    html_math: true
    theme: journal
    highlight: tango
    code_folding: show
---

```{r setup, include=FALSE}
# Minimal setup - only load essential libraries
knitr::opts_chunk$set(
  comment = "#>",
  eval = TRUE,
  message = FALSE
)
```

A lot of extra functionality comes from non-standard packages. After installing them, you can load them from your installed packages (your library?) using the `library()` function:

```{r packages}
# Load only necessary packages
library(dplyr)    # For data manipulation
library(ggplot2)  # For plotting
library(afex)     # For ANOVAs
```

The packages above: dplyr and ggplot2, are used for data frame handling and the "pipe" operator (dplyr) and for creating figures (ggplot2). They are part of the "tidyverse", a set of packages that work well together. You can install and load the whole tidyverse like this:

```{r tidyverse}
# Install the tidyverse (if not already installed)
# install.packages("tidyverse")

# Load the tidyverse
# library(tidyverse)
```

We will not explore the tidyverse any further here. However, most modern resources on using R for academic purposes do rely on the tidyverse, for example the free online version of the excellent book "R for Data Science" [https://r4ds.had.co.nz/](https://r4ds.had.co.nz/).

The next bit of code creates an artificial data set to play with. Run it, but no need to understand this next bit of code (yet).

```{r data set}

# Create sample data for demonstrations
set.seed(123)

# Factorial design data (Condition x Trial)
n_subjects <- 10
n_conditions <- 2
n_trials <- 5

data <- data.frame(
  ID = rep(1:n_subjects, each = 10),
  condition = rep(c("A", "B"), each = n_trials, n_subjects),
  trial = rep(1:n_trials, 2*n_subjects)
)

data$RT <- NA
data$RT[which(data$condition == "A")] <- rgamma(n=n_trials*n_subjects,
                                                shape=3,
                                                scale=0.075) + 0.5
data$RT[which(data$condition == "B")] <- rgamma(n=n_trials*n_subjects,
                                                shape=3,
                                                scale=0.075) + 0.6
weights <- (sample(c(1:n_subjects)) - (n_subjects/2) - 0.5) * (0.3/n_subjects)
for (ID in 1:n_subjects) {
  data$RT[which(data$ID == ID)] <- data$RT[which(data$ID == ID)] + weights[ID]
}

for (condition in c("A", "B")) {
  div <- ifelse(condition == "A", 5, 6)
  for (trial in c(1:5)) {
    idx <- which(data$condition == condition & data$trial == trial)
    data$RT[idx] <- data$RT[idx] + (log(5)/div) - (log(trial)/div)
  }
}
write.csv(data, "data.csv", row.names = FALSE)

```

---

## 1. Working with Data from CSV Files

This section covers basic data manipulation operations: reading, binding, merging, aggregating, and saving.

### 1.1 Reading CSV Files

```{r read_csv_example}
# Read a CSV file (assuming you have data.csv in your working directory)
df <- read.csv("data.csv")

# View the first few rows
head(df)

# Check data types
str(df)
```

### 1.2 Binding Multiple Data Frames

Combine data from different sources or files.

```{r bind_dataframes}
# Create sample data frames
df1 <- data.frame(ID = 1:5, Value = c(10, 20, 30, 40, 50))
df2 <- data.frame(ID = 6:10, Value = c(60, 70, 80, 90, 100))

# Horizontal binding (side-by-side)
df_horizontal <- cbind(df1, df2)

df_horizontal

# Vertical binding (stacked)
df_vertical <- rbind(df1, df2)

df_vertical

# Using bind_rows() from dplyr
df_stacked <- bind_rows(df1, df2)

df_stacked
# bind_cols() works similarly
```

### 1.3 Merging and Indexing Data Frames

Combine data based on common columns (like IDs).

```{r merge_dataframes}

# Create sample data with IDs
df1 <- data.frame(ID = 1:5, Value1 = rnorm(5))
df2 <- data.frame(ID = 1:5, Value2 = rnorm(5))

# Merge operation (inner join - only matching IDs)
df_merged <- merge(df1, df2, by = "ID")

df_merged

# we can select a column by name (or by index):
df1$ID

# or by using indexing, where we specify which rows AND which columns we want:
df1[1:5, 1]  # first 5 rows, first column

# we can select rows by index (although sometimes also by name)
# and get the whole row by leaving the column index empty:
df1[3,]

df2[c(2:5),]

# Full outer merge (all IDs from both data frames)
df_full <- merge(df1[c(1:4),], df2[c(2:5),], by = "ID", all = TRUE)

df_full
```

You can also use indexing to remove unnecessary columns, or maybe to remove trials (rows) with outliers, or errors.

### 1.4 Aggregating Data

Summarize data by grouping variables.

```{r aggregate_data}
# Sample data with grouping variable
data <- read.csv('data.csv')

# Using base R aggregate()
agg_base <- aggregate(RT ~ trial + condition, data = data, FUN = mean)

# now we have 8 rows: 2 groups x 4 trials:
agg_base

# the value of Value for Group1 and Trial 1
# is the mean of all Value entries in data
# that have Group == "Group1" and Trial == 1:

# Using dplyr group_by() and summary()
agg_dplyr <- data %>%
  group_by(trial, condition) %>%
  summarise(Mean = mean(RT, na.rm = TRUE))

agg_dplyr

# quick function to calculate confidence intervals for a vector of values
# these need to be normally distributed (if not, use bootstrapping instead)
CI <- function(X, conf=0.95) {
  m <- mean(X, na.rm = TRUE)
  s <- sd(X, na.rm = TRUE)
  n <- sum(!is.na(X))
  z <- qt((1-conf)/2, df=n-1, lower.tail=FALSE)
  hw <- z * (s / sqrt(n))
  return(c(m - hw, m + hw))
}

# Using dplyr group_by() and summarise() with multiple statistics
agg_comprehensive <- data %>%
  group_by(trial, condition) %>%
  summarise(
    Mean = mean(RT, na.rm = TRUE),
    SD = sd(RT, na.rm = TRUE),
    CI_lo = CI(RT, conf = 0.95)[1],
    CI_hi = CI(RT, conf = 0.95)[2],
    N = n()
  )

agg_comprehensive
```

### 1.5 Saving Data

Save your processed data to files.

```{r save_data}
# Save to CSV
write.csv(df, "processed_data.csv", row.names = FALSE)

# Save with additional parameters
write.csv(df, "data_with_special_chars.csv", 
          row.names = FALSE, na = "NA")

# Save as RDS (R-specific format, preserves all data types)
saveRDS(df, "data.rds")

# Verify files are created
file.exists("processed_data.csv")
```

---

## 2. Creating Standard Plots with ggplot2

This section demonstrates common plot types for visualizing experimental data.

### 2.1 Scatter Plots

```{r scatter_plot}
# Create sample data with different conditions
data <- read.csv('data.csv')

# Basic scatter plot with ggplot2
ggplot(data, aes(x = trial, y = RT)) +
  geom_point(aes(colour = condition), alpha = 0.7) +
  facet_wrap(~condition) +
  labs(title = "RT across trials by condition",
       x = "trial",
       y = "RT [s]")

# print(p_scatter)
```

### 2.2 Bar Charts for Comparing Conditions

```{r bar_chart}
# Create summary data for bar chart
# summary_data <- data %>%
#   group_by(condition) %>%
#   summarise(Mean = mean(Response, na.rm = TRUE),
#             SD = sd(Response, na.rm = TRUE)) %>%
#   mutate(Mean = as.numeric(Mean))

summary_data <- data %>%
  group_by(condition) %>%
  summarise(
    Mean = mean(RT, na.rm = TRUE),
    SD = sd(RT, na.rm = TRUE),
    CI_lo = CI(RT, conf = 0.95)[1],
    CI_hi = CI(RT, conf = 0.95)[2],
    N = n()
  )

# Bar chart with error bars
ggplot(summary_data, aes(x = condition, y = Mean)) +
  geom_bar(stat = "identity", fill = "steelblue") +
  geom_errorbar(aes(ymin = CI_lo, ymax = CI_hi),
                width = 0.2, size = 1) +
  labs(title = "Mean RT by condition",
       x = "condition",
       y = "mean RT")

# print(p_bar)
```

### 2.3 Line Graphs to Compare Conditions Over Time

```{r line_graph}
data <- read.csv('data.csv')

summary_data <- data %>%
  group_by(condition, trial) %>%
  summarise(
    Mean = mean(RT, na.rm = TRUE),
    SD = sd(RT, na.rm = TRUE),
    CI_lo = CI(RT, conf = 0.95)[1],
    CI_hi = CI(RT, conf = 0.95)[2],
    N = n()
  )

# Line plot showing trends over trials
ggplot(summary_data, aes(x = trial, y = Mean, colour = condition, group = condition)) +
  geom_line() +                                                      # plots as lines
  geom_ribbon(aes(ymin=CI_lo, ymax=CI_hi), linetype=0, alpha=0.1) +  # adds 95% CIs
  # geom_point() +                                                   # adds points
  labs(title = "RT over trials by condition",
       x = "trial",
       y = "RT [s]")  # adds labels to the axes


```

---

## 3. Simple Statistical Analysis

This section covers basic statistical tests using base R and the `afex` package.

### 3.1 One-Sample t-Test (Base R)

Test if the mean differs significantly from a specified value.

```{r one_sample_t_test}
# Sample data
data <- read.csv('data.csv')

# Perform one-sample t-test against a hypothesized mean
t_result <- t.test(data$RT, mu = 0.83)

# Display results
t_result

# Interpretation: If p < 0.05, reject the null hypothesis that the mean is 50
```

If the p-value is lower than 0.05 (the usual "alpha" level) we can say that there is an effect. This will happen if the "mu" we test the data against falls outside the confidence interval that is listed in the output.

### 3.2 Two-Sample t-Test (Base R)

Compare means between two independent groups.

```{r two_sample_t_test}
# Sample data with two conditions
data <- read.csv('data.csv')

# Two-sample t-test
t_result <- t.test(data$RT ~ data$condition)

# Display results
t_result

# This use Welch's t-test by default (more robust for unequal variances)
```

The listed confidence interval is now the confidence interval of the difference between the two samples (conditions). If this does not include 0, the p-value will be lower than 0.95.

### 3.3 ANOVA Using `afex` Package

The built-in ANOVA function does not use type III errors, which is more or less expected nowadays. Use the `afex` package to get modern ANOVA output.

```{r anova_afex}
# Example data
data <- read.csv('data.csv')

# Perform ANOVA using afex package
afex::aov_ez("ID", "RT", data, within = c("condition", "trial"))
```

In this case there is a 2x5 repeated measures ANOVA, with the factors "condition" (2 levels) and "trial" (5 levels). The output will show the main effects of condition (line 1) and trial (line 2), as well as their interaction (line 3).

If any of these effects are significant (p < 0.05), you could follow up with post-hoc tests to explore the differences further if there are more than 2 levels. For 2 levels we can see the effect directly in a figure (or by calculating the average within each of the levels).

This notebook demonstrates foundational R skills for undergraduate analysis.
All code is self-contained and can be adapted to various experimental designs.

Remember:
 - Always check your data first (head(), str(), summary())
 - Use meaningful variable names
 - Document your analysis steps clearly
 - Use appropriate statistical tests for your research questions

